---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reason 

This update is requested by Kurt Hornik on Nov 27, 2019, who found that the dependency `CALIBERrfimpute` is not on CRAN anymore. `mice 3.7.0` removes the dependency, an incorporates various modifications that accumulated over time.

## Checks

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash}
R CMD CHECK /Users/buurensv/Package/mice/mice_3.7.0.tar.gz
```


## Test environments

* local OS X install, 10.15.1, R 3.6.1
* win-builder, using `devtools::check_win_devel()`

Status: OK

## Downstream dependencies

I have run 

```{r eval=FALSE}
library("revdepcheck")
revdepcheck::revdep_reset()
revdep_check(num_workers = 3)
revdep_summary()
```

There were 57 reverse dependencies.
There were 5 packages that failed to check. None of these errors is mice-related.

## Failed to check (5)

|package                  |version  |error  |warning |note |
|:------------------------|:--------|:------|:-------|:----|
|[brms](failures.md#brms) |2.10.0   |__+1__ |        |1    |
|dynr                     |0.1.15-1 |1      |        |     |
|Hmisc                    |4.3-0    |1      |        |     |
|MissingDataGUI           |0.2-5    |1      |        |     |
|Replication              |0.1.1    |1      |        |     |

See <https://github.com/stefvanbuuren/mice/tree/master/revdep>

