---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reason 

This update is requested by Brian Ripley, who notes that the `DPpackage` 
will soon be archived. `mice 3.6.0` copies the `toenail` data from 
`DPpackage` and removes the package from `Suggests:`.

## Checks

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash}
R CMD CHECK /Users/buurensv/Package/mice/mice_3.6.0.tar.gz
```


## Test environments

* local OS X install, 10.14.5, R 3.6.0
* win-builder, using `devtools::check_win_devel()`

Status: OK

## Downstream dependencies

I have run 

```{r eval=FALSE}
library("revdepcheck")
revdepcheck::revdep_reset()
revdep_check(num_workers = 3)
revdep_summary()
```

There were 56 reverse dependencies.
There were 4 packages that failed to check. None of these errors is mice-related.

See <https://github.com/stefvanbuuren/mice/tree/master/revdep>

