---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## mice 3.16.0

## Reason

New submission in response to mail of CRAN team:

```
Dear maintainer,

Please see the problems shown on
<https://cran.r-project.org/web/checks/check_results_mice.html>.

Please correct before 2023-05-31 to safely retain your package on CRAN.

Packages in Suggests should be used conditionally: see 'Writing R Extensions'.
This needs to be corrected even if the missing package(s) become available.
It can be tested by checking with _R_CHECK_DEPENDS_ONLY_=true.

The CRAN Team
``` 

## Actions

- All issues noted in mail resolved
- Several changes and enhancements since `mice 3.15.0`

## Test environments

```{r}
R.Version()
```

## Checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()
...
WARNING   Requires (indirectly) orphaned package: 'ucminf'
```

Solving this is beyond my powers.

### Rhub

```{r eval=FALSE}
devtools::check_rhub()
```

Everything OK except for the orphaned package: 'ucminf'

### Local check

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK mice_3.16.0.tar.gz
```

Status: OK

## Downstream dependencies

### Overview

`mice` has 113 downstream dependencies

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(num_workers = 10)
```

```{r}
revdepcheck::revdep_summary()
```

### New issues

There is one new warning, for the `MatchThem` package:

```{r}
revdepcheck::revdep_details(revdep = "MatchThem")
```

### Actions taken

- I alerted the maintainer of the `MatchThem` package that the
  documentation for `cbind.mids()` was removed from the `mice` package
  to conform to CRAN guidelines.

- The novel issue for `bipd` is equivalent to the old one. I contacted the maintainer previously. Did not repeat it now.

- In the past, I alerted/contacted the maintainers of the packages with existing reverse dependencies issues. I did not repeat for this
  release of `mice`.
