---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Release summary

This is the resubmission of `mice` version 3.17.0. The previous submission was rejected due a problem with the downstream `autoReg` and `finalfit` packages. The problem was a change in the reporting of the confidence intervals. In the previous version confidence intervals were named `2.5 %` and `97.5 %`, a convention inherited from `stats::confint.default`. The update `mice 3.17.0` adopts the more flexible `broom` convention of naming these boundaries `conf.low` and `conf.high`, so that these names work for any probability level. This change caused the two downstream packages to fail. The issue has been fixed in this version of `mice` by duplicating the two conf columns, and naming them `2.5 %` and `97.5 %`. I will alert the maintainers of the `autoReg` and `finalfit` packages to this change, but for now, having the duplicate columns fixes the problem.

This release of `mice` adds new features, bug fixes, and documentation improvements. 
See <https://github.com/amices/mice/blob/master/NEWS.md>

## Test environments

```{r}
packageVersion("mice")
R.Version()$version.string
```

## Local checks

NOTE: Run in OSX terminal, not in Rstudio terminal.

```{bash eval=FALSE}
env _R_CHECK_DEPENDS_ONLY_=true R CMD check mice_3.17.0.tar.gz

Status: OK
```


```{r eval=FALSE}
devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))
...
Status: OK
```

## Remote checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()

Status: OK
```

### Rhub

```{r eval=FALSE}
rhub::rhub_doctor()
rhub::rhub_check()
```

Status: OK <>

For details: <https://github.com/amices/mice/actions>

## Downstream dependencies

### Code

Note: Inactivate `credentials::set_github_pat()`

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(pkg = ".", num_workers = 12, quiet = FALSE)
```

```{r}
revdepcheck::revdep_summary()
```

### revdepcheck results

We checked 123 reverse dependencies (118 from CRAN + 5 from Bioconductor), comparing R CMD check results across CRAN and dev versions of this package.

 * We saw 1 new problems
 * We failed to check 3 packages

Issues with CRAN packages are summarised below.

### New problems
(This reports the first line of each new failure)

* bipd
  checking dependencies in R code ...sh: line 1: 58727 Segmentation fault: 11  R_DEFAULT_PACKAGES=NULL '/Library/Frameworks/R.framework/Resources/bin/R' --vanilla --no-echo 2>&1 < '/var/folders/5_/g85d42yj50b6lrjq4rzjzg8w0000gn/T//RtmpM38C1S/filee1cf761be8f7'

### Failed to check

* brms (NA)
* dynr (NA)
* rmsb (NA)

mice developer comment: These failures are unrelated to mice

See <https://github.com/amices/mice/tree/master/revdep> for additional details.
