---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## mice 3.15.0

New submission. 

## Reason

`mice 3.15.0` contains many changes and enhancements over `mice 3.14.0`

## Test environments

### Local

```{r}
R.Version()
```

### win-builder

### * Rhub

## Local check

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK mice_3.14.15.tar.gz
```

Status: OK

## win-builder

```{r eval=FALSE}
devtools::check_win_devel()
```

Status: 1 NOTE


## Rhub checks

```{r eval=FALSE}
devtools::check_rhub()
```

Results:

1. Debian Linux, R-devel, GCC ASAN/UBSAN: 
2. Windows Server 2022, R-devel, 64 bit: **Success**
3. Ubuntu Linux 20.04.1 LTS, R-release, GCC: 
4. Fedora Linux, R-devel, clang, gfortran: 


## Downstream dependencies

I have run 

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(num_workers = 10)
```

### `failures.md`

There is one old failure (`dynr`): `configure: error: gsl-config not found, is GSL installed?`
Not related to `mice`.


### `problems.md`

```
# CALIBERrfimpute

<details>

* Version: 1.0-5
* GitHub: NA
* Source code: https://github.com/cran/CALIBERrfimpute
* Date/Publication: 2021-05-05 09:00:04
* Number of recursive dependencies: 53

Run `revdep_details(, "CALIBERrfimpute")` for more info

</details>

## Newly broken

*   checking running R code from vignettes ...
      ‘simstudy_survival.Rnw’ using ‘UTF-8’... failed
     ERROR
    Errors in running code in vignettes:
    when running code in ‘simstudy_survival.Rnw’
      ...
      x2 & 0.0391  & 0.0254  & 0.0138  \\ 
      x3 & -0.0314  & -0.00854  & -0.0228  \\ 
       \hline
    \end{tabular}
    
    \vspace{1em}
    
      When sourcing ‘simstudy_survival.R’:
    Error: missing value where TRUE/FALSE needed
    Execution halted
```

Not sure whether it's related to `mice`, and seems relatively benign if it is. The package maintainer of `CALIBERrfimpute` is aware of the problem, and will look into the issue.
