---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Release summary

`mice 3.19.10` adds new features, bug fixes, and enhancements.
See <https://github.com/amices/mice/blob/master/NEWS.md>

## Test environments

```{r}
packageVersion("mice")
R.Version()$version.string
```

## Local checks

NOTE: Run in OSX terminal, not in Rstudio terminal.

```{bash eval=FALSE}
env _R_CHECK_DEPENDS_ONLY_=true 
R CMD check mice_3.19.10.tar.gz

Status: OK
```


```{r eval=FALSE}
devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))
...
Status: OK
```

## Remote checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()

Upload error: timeout
```


### Platform checks

We checked the following platforms:

```{r eval=FALSE}
macos-latest (release)
windows-latest (release)
ubuntu-latest (devel)
ubuntu-latest (release)
ubuntu-latest (oldrel)
```

Status: OK

For details: <https://github.com/amices/mice/actions>

## Downstream dependencies

### Code

Note: Inactivate `credentials::set_github_pat()`

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(pkg = ".", num_workers = 12, quiet = FALSE)
```

```{r}
# revdepcheck::revdep_summary()
```


## revdepcheck results

We checked 177 reverse dependencies (171 from CRAN + 6 from Bioconductor), comparing R CMD check results across CRAN and dev versions of this package.

 * We saw 5 new problems
 * We failed to check 4 packages

Of the 5 new problems, only `autoReg` is caused by this release: it reads
the confidence interval columns from `summary(pool(fit), conf.int = TRUE)`
by their old `confint()`-style names (`` `2.5 %` ``/`` `97.5 %` ``), which
this version renames to `conf.low`/`conf.high` (broom convention). We have
contacted the maintainer about updating `autoReg` accordingly. The
remaining new problems (`bipd`, `broom.mixed`, `pminternal`, `pre`) are
unrelated environment flakiness (segfaults also seen as "newly fixed" in
the same run, or local port conflicts from parallel checking) and not
caused by changes in this package.

See <https://github.com/amices/mice/tree/master/revdep> for details.
