---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Package built by

```{r eval = FALSE}
library("devtools")
build_vignettes()
build()
```

`"/Users/buurensv/Package/mice/mice_3.0.0.tar.gz"`

## Test environments

* local OS X install, 10.13.4, R 3.5.0
* win-builder, using `devtools::build_win()`

There was no ERROR, WARNING or NOTE.

## R CMD check results

local checks of tarball fails:

`$ R CMD CHECK mice_3.0.0.tar.gz`

yields: 
`Error: package ‘Rcpp’ was installed by an R version with different internals; it needs to be reinstalled for use with this R version
Execution halted`

(I have installed `Rccp 0.12.17`)

`Error: package or namespace load failed for ‘lattice’:
 package ‘lattice’ was installed by an R version with different internals; it needs to be reinstalled for use with this R version`

(I have installed `lattice 0.20-35`)

Reinstalling does not help (`warning: package ‘Rccp’ is not available
(for R version 3.5.0)`). I cannot repair this problem. I have assumed
this is a temporary flitch that we may safely ignore for now.

## Downstream dependencies

I have run 

```{r eval=FALSE}
revdep_check()
revdep_check_save_summary()
revdep_check_print_problems()
```

There were 48 reverse dependencies.

Packages with problems + actions

|package         |version  | errors| warnings| notes| My actions
|:---------------|:--------|------:|--------:|-----:|:---------------------------------
|brms            |2.3.0    |      1|        0|     2| no error in brms 2.3.1 (SOLVED)
|CALIBERrfimpute |0.1-6    |      0|        1|     1| showtable() in simstudy_survival defunct - author noted
|codebook        |0.5.8    |      1|        0|     1| solved by change in md.pattern()
|dlookr          |0.3.0    |      1|        0|     0| author noted, not related to mice (SOLVED)
|dynr            |0.1.12-5 |      1|        0|     0| author noted, needs additional local installs, not further checked
|HardyWeinberg   |1.5.9    |      0|        1|     0| asked author whether error could be correct
|Hmisc           |4.1-1    |      1|        0|     0| installed by hand
|hot.deck        |1.1      |      0|        1|     0| warning unrelated to mice
|HSAUR3          |1.0-8    |      0|        1|     0| solved by reverting print.mira, broom naming, author noted 
|JointAI         |0.1.0    |      1|        0|     0| depends on external JAGS, not further checked
|logistf         |1.22     |      1|        0|     1| pooling code not compatible - author noted
|miceadds        |2.11-80  |      1|        0|     0| https://github.com/alexanderrobitzsch/miceadds/issues/6, will be solved in next release
|miceFast        |0.2.3    |      1|        0|     0| C compilation errors
|miceMNAR        |1.0      |      1|        0|     0| incompatible mice.impute.hecknorm(), author noted
|MissingDataGUI  |0.2-5    |      1|        0|     0| 'gWidgetsRGtk2', 'cairoDevice' not available
|Qtools          |1.3      |      1|        0|     0| fortran compilation errors, author noted
|rattle          |5.1.0    |      1|        0|     0| 'cairoDevice' not available
|sjmisc          |2.7.2    |      1|        0|     0| no errors found in sjmisc 2.7.2 (SOLVED)
|weightTAPSPACK  |0.1      |      1|        0|     0| compilation error in Rglpk

See https://github.com/stefvanbuuren/mice/blob/master/revdep/problems.md
