---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `mice 3.9.0`, new submission

## Reason 

This update `mice 3.9.0` adds new functionality, reduces dependencies and provides solutions for issues found in `mice 3.8.0` and before.

## Checks

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK /Users/buurensv/Package/mice/mice_3.9.0.tar.gz
```

This fails!? (although interactive building in Rstudio works fine)

```
* installing *source* package ‘mice’ ...
** using staged installation
** libs
clang++ -mmacosx-version-min=10.13 -std=gnu++11 -I"/Library/Frameworks/R.framework/Resources/include" -DNDEBUG  -I'/Users/buurensv/Library/R/3.6/library/Rcpp/include' -I/usr/local/include   -fPIC  -Wall -g -O2  -c RcppExports.cpp -o RcppExports.o
clang++ -mmacosx-version-min=10.13 -std=gnu++11 -I"/Library/Frameworks/R.framework/Resources/include" -DNDEBUG  -I'/Users/buurensv/Library/R/3.6/library/Rcpp/include' -I/usr/local/include   -fPIC  -Wall -g -O2  -c match.cpp -o match.o
Error: package ‘Rcpp’ was installed before R 4.0.0: please re-install it
Execution halted
clang++ -mmacosx-version-min=10.13 -std=gnu++11 -dynamiclib -Wl,-headerpad_max_install_names -undefined dynamic_lookup -single_module -multiply_defined suppress -L/Library/Frameworks/R.framework/Resources/lib -L/usr/local/lib -o mice.so RcppExports.o match.o -F/Library/Frameworks/R.framework/.. -framework R -Wl,-framework -Wl,CoreFoundation
Error: package ‘Rcpp’ was installed before R 4.0.0: please re-install it
Execution halted
installing to /Users/buurensv/Dropbox/Package/mice/mice/mice.Rcheck/00LOCK-mice/00new/mice/libs
** R
** data
*** moving datasets to lazyload DB
** inst
** byte-compile and prepare package for lazy loading
Error: package ‘broom’ was installed before R 4.0.0: please re-install it
Execution halted
ERROR: lazy loading failed for package ‘mice’
* removing ‘/Users/buurensv/Dropbox/Package/mice/mice/mice.Rcheck/mice’
```

I do not understand this message. After upgrading the `R 4.0.0`, I refreshed all packages, as required. Re-installing `Rcpp` manually did not solve the problem.

## Test environments

* local OS X install, 10.15.4, R 4.0.0
* win-builder, using `devtools::check_win_devel()`

Status:

```

```


```{r eval=FALSE}
check_rhub()
```

The upload failed.

```
─  Uploading package
Error in query("SUBMIT PACKAGE", data = list(email = unbox(email), token = unbox(email_get_token(email)),  : 
  Email address not validated
> 
```

## Downstream dependencies

I have run 

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(num_workers = 3)
revdep_summary()
```

### `problems.md`

```
*Wow, no problems at all. :)*
```

### README.md 

# Platform

|field    |value                        |
|:--------|:----------------------------|
|version  |R version 4.0.0 (2020-04-24) |
|os       |macOS Catalina 10.15.4       |
|system   |x86_64, darwin17.0           |
|ui       |RStudio                      |
|language |(EN)                         |
|collate  |en_US.UTF-8                  |
|ctype    |en_US.UTF-8                  |
|tz       |Europe/Amsterdam             |
|date     |2020-05-14                   |

# Dependencies

|package    |old      |new      |Δ  |
|:----------|:--------|:--------|:--|
|mice       |3.8.0    |3.9.0    |*  |
|assertthat |0.2.1    |0.2.1    |   |
|backports  |1.1.7    |1.1.7    |   |
|BH         |1.72.0-3 |1.72.0-3 |   |
|broom      |0.5.6    |0.5.6    |   |
|cli        |2.0.2    |2.0.2    |   |
|crayon     |1.3.4    |1.3.4    |   |
|digest     |0.6.25   |0.6.25   |   |
|dplyr      |0.8.5    |0.8.5    |   |
|ellipsis   |0.3.0    |0.3.0    |   |
|fansi      |0.4.1    |0.4.1    |   |
|generics   |0.0.2    |0.0.2    |   |
|glue       |1.4.1    |1.4.1    |   |
|lifecycle  |0.2.0    |0.2.0    |   |
|magrittr   |1.5      |1.5      |   |
|pillar     |1.4.4    |1.4.4    |   |
|pkgconfig  |2.0.3    |2.0.3    |   |
|plogr      |0.2.0    |0.2.0    |   |
|plyr       |1.8.6    |1.8.6    |   |
|purrr      |0.3.4    |0.3.4    |   |
|R6         |2.4.1    |2.4.1    |   |
|Rcpp       |1.0.4.6  |1.0.4.6  |   |
|reshape2   |1.4.4    |1.4.4    |   |
|rlang      |0.4.6    |0.4.6    |   |
|stringi    |1.4.6    |1.4.6    |   |
|stringr    |1.4.0    |1.4.0    |   |
|tibble     |3.0.1    |3.0.1    |   |
|tidyr      |1.0.3    |1.0.3    |   |
|tidyselect |1.1.0    |1.1.0    |   |
|utf8       |1.1.4    |1.1.4    |   |
|vctrs      |0.3.0    |0.3.0    |   |

## Revdeps

### Failed to check (4)

|package        |version   |error |warning |note |
|:--------------|:---------|:-----|:-------|:----|
|brms           |2.12.0    |1     |        |1    |
|dynr           |0.1.15-25 |1     |        |     |
|MissingDataGUI |0.2-5     |1     |        |     |
|Replication    |0.1.2     |1     |        |     |

See 
