---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## mice 3.17.0

## Overview

This submission of `mice` adds new features, bug fixes, and documentation improvements. 
See <https://github.com/amices/mice/blob/master/NEWS.md>

The package has been tested on the CRAN incoming checks and on the `win-builder` and `rhub` platforms. The package has 132 downstream dependencies. I found one issue for which I contacted the maintainer of the `autoreg` package.

## Test environments

```{r}
R.Version()$version.string
```

## Checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()

Status: OK
```


### Rhub

```{r eval=FALSE}
rhub::rhub_doctor()
rhub::rhub_check()
```

Status: OK <https://github.com/amices/mice/actions/runs/11974214695>


### Local checks

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK mice_3.17.0.tar.gz

Status: OK
```


### Local check using `_R_CHECK_DEPENDS_ONLY_=true` flag. 

NOTE: Run in OSX terminal, not in Rstudio terminal.

```{bash eval=FALSE}
env _R_CHECK_DEPENDS_ONLY_=true R CMD check mice_3.17.0.tar.gz

Status: OK
```


```{r eval=FALSE}
devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))

...
Status: 1 NOTE

── R CMD check results ─────────────────────────────────────────────────────────────────────── mice 3.17.0 ────
Duration: 1m 10.9s

❯ checking for future file timestamps ... NOTE
  unable to verify current time

0 errors ✔ | 0 warnings ✔ | 1 note ✖
```


## Downstream dependencies

### Overview

`mice` has 132 downstream dependencies

```{r eval=FALSE}
# NOTE: Temporarily remove credentials line from .Rprofile
library(revdepcheck)
revdep_reset()
revdep_check(pkg = ".", num_workers = 10, quiet = FALSE)
```

```{r}
revdepcheck::revdep_summary()
```

# Revdeps

## Failed to check (5)

|package     |version    |error |warning |note |                    |
|:-----------|:----------|:-----|:-------|:----| ------------------ | 
|brms        |2.22.0     |1     |        |2    | Unrelated to mice  |
|dynr        |0.1.16-105 |1     |        |     | Unrelated to mice  |
|pguIMP      |0.0.0.3    |1     |        |     | Unrelated to mice  |
|Replication |0.1.2      |1     |        |     | Unrelated to mice  |
|rmsb        |1.1-1      |1     |        |1    | Unrelated to mice  |

## New problems (4)

|package  |version |error  |warning |note      |                   |
|:--------|:-------|:------|:-------|:---------|:------------------|
|[autoReg](problems.md#autoreg)|0.3.3   |__+2__ |        |          |  Informed the maintainer, asked for repair |
|[bipd](problems.md#bipd)|0.3     |1      |        |-1 __+1__ |  Does not run because of different CPU's |
|[finalfit](problems.md#finalfit)|1.0.8   |__+2__ |        |1         | Same error as autoreg, possibly related |
|[pre](problems.md#pre)|1.0.7   |__+1__ |        |          | Unrelated to mice |

See <https://github.com/amices/mice/tree/master/revdep> for details.
