---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Release summary

This release of `mice` adds new features, bug fixes, and documentation improvements. 
See <https://github.com/amices/mice/blob/master/NEWS.md>

## Test environments

```{r}
packageVersion("mice")
R.Version()$version.string
```

## Local checks

NOTE: Run in OSX terminal, not in Rstudio terminal.

```{bash eval=FALSE}
env _R_CHECK_DEPENDS_ONLY_=true R CMD check mice_3.17.0.tar.gz

Status: OK
```


```{r eval=FALSE}
devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))
...
Status: OK
```

## Remote checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()

Status: OK
```

### Rhub

```{r eval=FALSE}
rhub::rhub_doctor()
rhub::rhub_check()
```

Status: OK <>

For details: <https://github.com/amices/mice/actions>

## Downstream dependencies

### Code

Note: Inactivate `credentials::set_github_pat()`

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(pkg = ".", num_workers = 12, quiet = FALSE)
```

```{r}
revdepcheck::revdep_summary()
```

### Summary

We checked 123 reverse dependencies (118 from CRAN + 5 from Bioconductor), comparing R CMD check results across CRAN and dev versions of this package.

 * We saw 3 new problems
 * We failed to check 3 packages

Issues with CRAN packages are summarised below.

### New problems (3)

|package  |version |error  |warning |note      |
|:--------|:-------|:------|:-------|:---------|
|[autoReg](problems.md#autoreg)|0.3.3   |__+2__ |        |          |
|[bipd](problems.md#bipd)|0.3     |1      |        |-1 __+1__ |
|[finalfit](problems.md#finalfit)|1.0.8   |__+2__ |        |1         |

(This reports the first line of each new failure)

* autoReg
  checking examples ... ERROR
  checking running R code from vignettes ...

* bipd
  checking dependencies in R code ...sh: line 1: 24340 Segmentation fault: 11  R_DEFAULT_PACKAGES=NULL '/Library/Frameworks/R.framework/Resources/bin/R' --vanilla --no-echo 2>&1 < '/var/folders/5_/g85d42yj50b6lrjq4rzjzg8w0000gn/T//RtmpazqvFP/file5c206c170a3d'

* finalfit
  checking examples ... ERROR
  checking running R code from vignettes ...

mice developer comment: These new problems are unrelated to mice

### Failed to check (3)

|package |version    |error |warning |note |
|:-------|:----------|:-----|:-------|:----|
|brms    |2.22.0     |1     |        |2    |
|dynr    |0.1.16-105 |1     |        |     |
|rmsb    |1.1-1      |1     |        |1    |

* brms (NA)
* dynr (NA)
* rmsb (NA)

mice developer comment: These failures are unrelated to mice

See <https://github.com/amices/mice/tree/master/revdep> for additional details.
