---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## mice 3.16.0

## Overview

This submission 

- is a third resubmission, using `devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))` to mimmick CRAN incoming checks
- is a second resubmission, with `\donttest` changed in `\dontrun` to evade `_R_CHECK_DEPENDS_ONLY_=true` errors;
- is a resubmission of `mice 3.16.0` after the orphaned `ucminf` CRAN package got a new maintainer;
- is tested with the `_R_CHECK_DEPENDS_ONLY_=true` flag, as requested;
- solves the problems at `https://cran.r-project.org/web/checks/check_results_mice.html`;
- contains new features and bug fixes, as described in the `NEWS.md`.

## Test environments

```{r}
R.Version()
```

## Checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()

Status: OK
```


### Rhub

```{r eval=FALSE}
devtools::check_rhub()
```

#### Windows

```
── mice 3.16.0: NOTE

  Build ID:   mice_3.16.0.tar.gz-018f40fa4cdc492d8c9f43b7e5f5e0e7
  Platform:   Windows Server 2022, R-devel, 64 bit
  Submitted:  11m 27s ago
  Build time: 11m 25s

❯ checking HTML version of manual ... [27s] NOTE
  Skipping checking math rendering: package 'V8' unavailable

❯ checking for non-standard things in the check directory ... NOTE
  Found the following files/directories:
    ''NULL''

❯ checking for detritus in the temp directory ... NOTE
  Found the following files/directories:
    'lastMiKTeXException'

0 errors ✔ | 0 warnings ✔ | 3 notes ✖
```

I believe these NOTES are benign.

#### Fedora Linux

```
Status: succes
```

#### Ubuntu 20.04

```
Status: succes
```

#### Debian

```
FAILURE 

Running `R CMD build`...
* checking for file ‘/tmp/Rtmp5Lhzhn/remotes13171d81356/mice/DESCRIPTION’ ... OK
* preparing ‘mice’:
* checking DESCRIPTION meta-information ... OK
* cleaning src
* checking for LF line-endings in source and make files and shell scripts
* checking for empty or unneeded directories
* building ‘mice_3.16.0.tar.gz’
Installing package into ‘/home/docker/R’
(as ‘lib’ is unspecified)
Error : Bioconductor does not yet build and check packages for R version 4.4; see
  https://bioconductor.org/install
ERROR: dependency ‘mitml’ is not available for package ‘mice’
* removing ‘/home/docker/R/mice’
> There were 20 warnings (use warnings() to see them)

> 
Error : Bioconductor does not yet build and check packages for R version 4.4; see
  https://bioconductor.org/install
> library(mice)
Error in library(mice) : there is no package called ‘mice’
Execution halted
Build step 'Execute shell' marked build as failure
Pinging https://builder.r-hub.io/build/FAILURE/mice_3.16.0.tar.gz-82dc7cda08814b25ac48be4b5c9c940b/2023-06-01T11:37:52Z
{"status":"ok"}
Finished: FAILURE
```

I believe I cannot do something to make this problem disappear.

### Local checks

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK mice_3.16.0.tar.gz

Status: OK
```


### Local check using `_R_CHECK_DEPENDS_ONLY_=true` flag. 

NOTE: Run in OSX terminal, not in Rstudio terminal.

```{bash eval=FALSE}
env _R_CHECK_DEPENDS_ONLY_=true R CMD check mice_3.16.0.tar.gz

Status: OK
```


```{r eval=FALSE}
devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))

...
   Status: OK
   
── R CMD check results ─────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────── mice 3.16.0 ────
Duration: 1m 13s

0 errors ✔ | 0 warnings ✔ | 0 notes ✔
```


## Downstream dependencies

### Overview

`mice` has 114 downstream dependencies

```{r eval=FALSE}
# NOTE: Temporarily remove credentials line from .Rprofile
library(revdepcheck)
revdep_reset()
revdep_check(pkg = ".", num_workers = 10, quiet = FALSE)
```

```{r}
revdepcheck::revdep_summary()
```

### New issues

#### `MatchThem`

There is one new warning, for the `MatchThem` package:

```{r}
revdepcheck::revdep_details(revdep = "MatchThem")
```

- I alerted the maintainer of the `MatchThem` package that the
  documentation for `cbind.mids()` was removed from the `mice` package
  to conform to CRAN guidelines. The maintainer responded this will be fixed
  in the next release.

#### `bipd`

```{r}
revdepcheck::revdep_details(revdep = "bipd")
```

- The novel NOTE for `bipd` is equivalent to the old one. I contacted the maintainer previously. Did not repeat it now.

#### `pre`

- There is a new error generated by `test_pre_misc.R` of the `pre` package (`mice` is on `pre` Suggests):

```{r}
revdepcheck::revdep_details(revdep = "pre")
```

- I tried reproducing the error but stopped my attempt due to the complex structure 
  of the testing script. I contacted the `pre` maintainer by means of 
  issue <https://github.com/marjoleinF/pre/issues/30>.

