---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## mice 3.12.0

New submission. 

## Reason

`mice 3.12.0` implements important new features and contains bug fixes.

## Test environments

* local OS X install, 10.15.7, R 4.0.3
* win-builder
* Rhub

## Local check

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK mice_3.12.0.tar.gz
```


Status: OK

## win-builder

```r
devtools::check_win_devel()
```

Status: OK


## Rhub checks

```{r eval=FALSE}
check_rhub()
```

Results:

1. Debian Linux, R-devel, GCC ASAN/UBSAN: <https://builder.r-hub.io/status/mice_3.12.0.tar.gz-0f353a5af0994e03a4d81cfe59b64208>: Error, possibly related to an installing problem of miceadds package
2. Windows Server 2008 R2 SP1, R-devel, 32/64 bit: <https://builder.r-hub.io/status/mice_3.12.0.tar.gz-7b498c42ea2e4649890a59cae55aceb4>: Success
3. Ubuntu Linux 16.04 LTS, R-release, GCC: <https://builder.r-hub.io/status/mice_3.12.0.tar.gz-c2ce58219fd54312ab8ea65cddafcf8f>: Success
4. Fedora Linux, R-devel, clang, gfortran: <https://builder.r-hub.io/status/mice_3.12.0.tar.gz-b7d7fa09d5a54f97a0d475bb0f61339b>: Error, likely to be unrelated to mice


## Downstream dependencies

I have run 

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(num_workers = 10)
```

### `failures.md`

There are one new failure (`brms`) and one old (`dynr`). Both failures are caused by 
installation problems on the test servers. These failure do not seem to be related to `mice`.

### `problems.md`

There are new warnings in several packages because `mice 3.12.0` exports a new 
`filter.mids()` method for the `dplyr::filter()` generic. For example, we 
get `Warning: replacing previous import ‘mice::filter’ by ‘stats::filter’ when loading ‘BaM’`.
Sometimes we get the reverse message. I noted the package authors the upcoming 
change in mice on Oct 29. Affected packages: 
`BaM`, `genpathmox`, `idem`, `misaem`, `Qtools`, `TestDataImputation`

Here's the mail:

--- 

R packages: BaM, genpathmox, idem, misaem, Qtools, TestDataImputation

Dear package maintainer, 

A reversed dependency check revealed a new warning in your package that results from the upcoming release mice 3.12.0. The new release exports a new `filter.mids()` method for the `dplyr::filter()` generic. 

I  get: `Warning: replacing previous import ‘mice::filter’ by ‘stats::filter’ when loading ‘BaM’` and others. For Qtools I get the reverse: Warning: replacing previous import ‘stats::filter’ by ‘mice::filter’ when loading ‘Qtools’.

The warning is probably benign, but please check in your package if the introduction of `mice::filter.mids()` would inadvertent change a call to `stats::filter()` to `dplyr::filter()` or vice versa. If this happens, please update your import directive. 

I intend to submit to CRAN in about one week. Hope this helps to evade surprises when mice 3.12.0 hits CRAN.

Best wishes, Stef

---

Update Nov 9: The package maintainers of `BaM` and `Qtools` told me that the planned mice update did not affect the validity of their packages. Did not see a reaction from other maintainers.

A second problem concerned to `miceadds` package. This package uses 
`mice::.pmm.match`, which I had removed because I thought nobody used it 
anymore. I will put it back in.

Update Nov 9: `mice::.pmm.match` is now again part of the `mice` package

The `idem` package yields two notes. Neither of these is related to `mice`.

### README.md 

```
# Platform

|field    |value                        |
|:--------|:----------------------------|
|version  |R version 4.0.2 (2020-06-22) |
|os       |macOS Catalina 10.15.7       |
|system   |x86_64, darwin17.0           |
|ui       |RStudio                      |
|language |(EN)                         |
|collate  |en_US.UTF-8                  |
|ctype    |en_US.UTF-8                  |
|tz       |Europe/Amsterdam             |
|date     |2020-10-29                   |

# Dependencies

|package    |old    |new    |Δ  |
|:----------|:------|:------|:--|
|mice       |3.11.0 |3.12.0 |*  |
|assertthat |0.2.1  |0.2.1  |   |
|backports  |1.1.10 |1.1.10 |   |
|broom      |0.7.2  |0.7.2  |   |
|cli        |2.1.0  |2.1.0  |   |
|cpp11      |0.2.3  |0.2.3  |   |
|crayon     |1.3.4  |1.3.4  |   |
|digest     |0.6.27 |0.6.27 |   |
|dplyr      |1.0.2  |1.0.2  |   |
|ellipsis   |0.3.1  |0.3.1  |   |
|fansi      |0.4.1  |0.4.1  |   |
|generics   |0.0.2  |0.0.2  |   |
|glue       |1.4.2  |1.4.2  |   |
|lifecycle  |0.2.0  |0.2.0  |   |
|magrittr   |1.5    |1.5    |   |
|pillar     |1.4.6  |1.4.6  |   |
|pkgconfig  |2.0.3  |2.0.3  |   |
|purrr      |0.3.4  |0.3.4  |   |
|R6         |2.5.0  |2.5.0  |   |
|Rcpp       |1.0.5  |1.0.5  |   |
|rlang      |0.4.8  |0.4.8  |   |
|stringi    |1.5.3  |1.5.3  |   |
|stringr    |1.4.0  |1.4.0  |   |
|tibble     |3.0.4  |3.0.4  |   |
|tidyr      |1.1.2  |1.1.2  |   |
|tidyselect |1.1.0  |1.1.0  |   |
|utf8       |1.1.4  |1.1.4  |   |
|vctrs      |0.3.4  |0.3.4  |   |

# Revdeps

## Failed to check (2)

|package |version   |error |warning |note |
|:-------|:---------|:-----|:-------|:----|
|brms    |?         |      |        |     |
|dynr    |0.1.15-25 |1     |        |     |

## New problems (7)

|package                                              |version |error |warning |note     |
|:----------------------------------------------------|:-------|:-----|:-------|:--------|
|[BaM](problems.md#bam)                               |1.0.1   |      |__+1__  |         |
|[genpathmox](problems.md#genpathmox)                 |0.4     |      |__+1__  |         |
|[idem](problems.md#idem)                             |5.0     |      |__+1__  |2        |
|[miceadds](problems.md#miceadds)                     |3.10-28 |      |        |2 __+1__ |
|[misaem](problems.md#misaem)                         |1.0.0   |      |__+1__  |         |
|[Qtools](problems.md#qtools)                         |1.5.2   |      |__+1__  |         |
|[TestDataImputation](problems.md#testdataimputation) |1.1     |      |__+1__  |         |
```

This package requires additional software to be installed. See <https://github.com/amices/mice/blob/master/revdep/failures.md> for details.
