---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Release summary

`mice 3.19.0` adds new features, bug fixes, and enhancements. 
See <https://github.com/amices/mice/blob/master/NEWS.md>

## Test environments

```{r}
packageVersion("mice")
R.Version()$version.string
```

## Local checks

NOTE: Run in OSX terminal, not in Rstudio terminal.

```{bash eval=FALSE}
env _R_CHECK_DEPENDS_ONLY_=true 
R CMD check mice_3.19.0.tar.gz

Status: OK
```


```{r eval=FALSE}
devtools::check(env_vars = c('_R_CHECK_DEPENDS_ONLY_' = "true"))
...
Status: OK
```

## Remote checks

### win-builder

```{r eval=FALSE}
devtools::check_win_devel()

Upload error: timeout
```


### Platform checks

We checked the following platforms:

```{r eval=FALSE}
macos-latest (release)
windows-latest (release)
ubuntu-latest (devel)
ubuntu-latest (release)
ubuntu-latest (oldrel)
```

Status: OK

For details: <https://github.com/amices/mice/actions>

## Downstream dependencies

### Code

Note: Inactivate `credentials::set_github_pat()`

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(pkg = ".", num_workers = 12, quiet = FALSE)
```

```{r}
# revdepcheck::revdep_summary()
```


## revdepcheck results

We checked 148 reverse dependencies (140 from CRAN + 8 from Bioconductor), comparing R CMD check results across CRAN and dev versions of this package.

 * We saw 1 new problems
 * We failed to check 5 packages

See <https://github.com/amices/mice/tree/master/revdep> for details.
