---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## mice 3.10
New submission

## Reason 

`mice 3.10.0` adds new functions `tidy.mipo()` and `glance.mipo()`, and solves a problem caused by separating `broom` and `broom.mixed`.

## Test environments

* local OS X install, 10.15.4, R 4.0.2
* win-builder
* Rhub

## Local check

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK /Users/buurensv/Package/mice/mice_3.10.0.tar.gz
```


Status: OK

## win-builder

```r
devtools::check_win_devel()
```

Status: OK


## Rhub checks

```{r eval=FALSE}
check_rhub()
```

```
Build ID:	mice_3.10.0.tar.gz-ad5f8a9c8d27d823a9c47247b8a27a39
Platform:	Debian Linux, R-devel, GCC
Submitted:	58 minutes 23 seconds ago
Build time:	57 minutes 56.7 seconds

Status: succes
```

```
Build ID:	mice_3.10.0.tar.gz-c5935e091347f67ee83d2577042c897d
Platform:	Debian Linux, R-devel, GCC
Submitted:	28 minutes 51.1 seconds ago
Build time:	28 minutes 49 seconds

Status: succes
```


## Downstream dependencies

I have run 

```{r eval=FALSE}
library(revdepcheck)
revdep_reset()
revdep_check(num_workers = 3)
revdep_summary()
```

### `problems.md`

```
*Wow, no problems at all. :)*
```

### README.md 

```
# Platform

|field    |value                        |
|:--------|:----------------------------|
|version  |R version 4.0.2 (2020-06-22) |
|os       |macOS Catalina 10.15.5       |
|system   |x86_64, darwin17.0           |
|ui       |RStudio                      |
|language |(EN)                         |
|collate  |en_US.UTF-8                  |
|ctype    |en_US.UTF-8                  |
|tz       |Europe/Amsterdam             |
|date     |2020-07-13                   |

# Dependencies

|package    |old    |new    |Δ  |
|:----------|:------|:------|:--|
|mice       |3.9.0  |3.10.0 |*  |
|assertthat |0.2.1  |0.2.1  |   |
|backports  |1.1.8  |1.1.8  |   |
|broom      |0.7.0  |0.7.0  |   |
|cli        |2.0.2  |2.0.2  |   |
|crayon     |1.3.4  |1.3.4  |   |
|digest     |0.6.25 |0.6.25 |   |
|dplyr      |1.0.0  |1.0.0  |   |
|ellipsis   |0.3.1  |0.3.1  |   |
|fansi      |0.4.1  |0.4.1  |   |
|generics   |0.0.2  |0.0.2  |   |
|glue       |1.4.1  |1.4.1  |   |
|lifecycle  |0.2.0  |0.2.0  |   |
|magrittr   |1.5    |1.5    |   |
|pillar     |1.4.6  |1.4.6  |   |
|pkgconfig  |2.0.3  |2.0.3  |   |
|purrr      |0.3.4  |0.3.4  |   |
|R6         |2.4.1  |2.4.1  |   |
|Rcpp       |1.0.5  |1.0.5  |   |
|rlang      |0.4.7  |0.4.7  |   |
|stringi    |1.4.6  |1.4.6  |   |
|stringr    |1.4.0  |1.4.0  |   |
|tibble     |3.0.3  |3.0.3  |   |
|tidyr      |1.1.0  |1.1.0  |   |
|tidyselect |1.1.0  |1.1.0  |   |
|utf8       |1.1.4  |1.1.4  |   |
|vctrs      |0.3.1  |0.3.1  |   |

# Revdeps

## Failed to check (2)

|package     |version   |error |warning |note |
|:-----------|:---------|:-----|:-------|:----|
|dynr        |0.1.15-25 |1     |        |     |
|Replication |0.1.2     |1     |        |     |
```

These package require additional software to be installed. See <https://github.com/stefvanbuuren/mice/blob/master/revdep/failures.md> for details.
