---
title: "mice: Multivariate Imputation by Chained Equations in R"
author: "Stef van Buuren, Karin Groothuis-Oudshoorn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: refs.bibtex
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, fig.height = 5, fig.width = 7, 
                      tidy = TRUE, fig.retina = 2, cache = TRUE)
options(digits = 3, width = 55)
set.seed(92622)
```

This page contains the R code from the article [mice: Multivariate Imputation by Chained Equations in R](https://www.jstatsoft.org/v45/i03/paper) published in the Journal of Statistical Software [@vanbuuren2011a]. For other examples using MICE see [MICE: Multivariate Imputation by Chained Equations](http://stefvanbuuren.github.io/mice/). 

The [source code](https://raw.githubusercontent.com/stefvanbuuren/mice/gh-pages/JSS.Rmd) was tested in R 3.4.2 using mice 2.44.

### Install packages needed to run code
```{r}
list.of.packages <- c("mice", "VIM", "mitools", "Zelig")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```


### 2.4 Simple example

#### chunk number 4: example1
```{r}
library("mice")
```


#### chunk number 5: example2 
```{r}
nhanes
```


#### chunk number 6: example3
```{r}
md.pattern(nhanes)
```


#### chunk number 7: example4
```{r}
p <- md.pairs(nhanes)
p
```


#### chunk number 9: marginplot
```{r}
VIM::marginplot(nhanes[, c("chl", "bmi")], col = mdc(1:2), 
  cex = 1.2, cex.lab = 1.2, cex.numbers = 1.3, pch = 19)
```


#### chunk number 10: example1
```{r}
imp <- mice(nhanes, seed = 23109)
```


#### chunk number 11: example2
```{r}
print(imp)
```


#### chunk number 12: example3
```{r}
imp$imp$bmi
```


#### chunk number 13: example4 
```{r}
complete(imp)
```


#### chunk number 14: stripplotcmd 
```{r}
stripplot(imp, pch = 20, cex = 1.2)
```


#### chunk number 16: xyplotcmd 
```{r}
xyplot(imp, bmi ~ chl | .imp, pch = 20, cex = 1.4)
```



#### chunk number 18: example6
```{r}
fit <- with(imp, lm(chl ~ age + bmi))
```



#### chunk number 19: example7
```{r}
print(pool(fit))
```



#### chunk number 20: example8
```{r}
round(summary(pool(fit)), 2)
```



#### chunk number 21: example9
```{r}
imp50 <- mice(nhanes, m = 50, seed = 23109, print = FALSE)
fit <- with(imp50, lm(chl ~ age + bmi))
```



#### chunk number 22: example10
```{r}
round(summary(pool(fit)), 2)
```


### 3.2 Univariate imputation methods

#### chunk number 23: models1 
```{r}
imp <- mice(nhanes, method = "norm", print = FALSE)
```


#### chunk number 24: models2 
```{r}
imp <- mice(nhanes, meth = c("", "norm", "pmm", "mean"), print = FALSE)
```



#### chunk number 25: models3
```{r}
str(nhanes2)
```



#### chunk number 26: models4 
```{r}
imp <- mice(nhanes2, me = c("polyreg", "pmm", "logreg", 
  "norm"), print = FALSE)
```



#### chunk number 27: models5 
```{r}
imp <- mice(nhanes2, meth = c("", "", "logreg", "norm"), print = FALSE)
```



#### chunk number 28: models6 
```{r}
mice(nhanes2, defaultMethod = c("norm", "logreg", "polyreg", 
  "polr"), print = FALSE)
```


### 3.3 Predictor selection

#### chunk number 29: predictor1
```{r}
imp <- mice(nhanes, print = FALSE)
imp$predictorMatrix
```



#### removing a predictor chunk number 30: predictor2
```{r}
pred <- imp$predictorMatrix
pred[, "bmi"] <- 0
pred
```



#### chunk number 31: predictor3
```{r}
imp <- mice(nhanes, pred = pred, print = FALSE)
```



#### skipping imputation chunk number 32: predictor4
```{r}
ini <- mice(nhanes2, maxit = 0, print = FALSE)
pred <- ini$pred
pred[, "bmi"] <- 0
meth <- ini$meth
meth["bmi"] <- ""
imp <- mice(nhanes2, meth = meth, pred = pred, print = FALSE)
imp$imp$bmi
```



#### intercept imputation chunk number 33: predictor5
```{r}
pred <- ini$pred
pred["bmi", ] <- 0
imp <- mice(nhanes2, pred = pred, print = FALSE, seed = 51162)
imp$imp$bmi
```


#### multilevel imputation chunk number 34:
```{r}
### multilevel1
popmis[1:3, ]
ini <- mice(popmis, maxit = 0, print = FALSE)
pred <- ini$pred
pred["popular", ] <- c(0, -2, 0, 2, 1, 2, 0)
imp <- mice(popmis, meth = c("", "", "2l.norm", "", 
  "", "", ""), pred = pred, maxit = 1, seed = 71152, print = FALSE)
```


#### chunk number 35: quickpred1
```{r}
round(cor(nhanes, use = "pair"), 3)
```



#### chunk number 36: quickpred2
```{r}
round(cor(y = nhanes, x = !is.na(nhanes), use = "pair"), 
  3)
```



#### chunk number 37: quickpred3
```{r}
p <- md.pairs(nhanes)
round(p$mr/(p$mr + p$mm), 3)
```



#### chunk number 38: quickpred4
```{r}
quickpred(nhanes)
```



#### chunk number 39: quickpred5 
```{r}
imp <- mice(nhanes, pred = quickpred(nhanes, minpuc = 0.25, 
  include = "age"), print = FALSE)
```


### 3.4 Passive imputation

#### Preserving, chunk number 40: transform1
```{r}
nhanes2.ext <- cbind(nhanes2, lchl = log(nhanes2$chl))
ini <- mice(nhanes2.ext, max = 0, print = FALSE)
meth <- ini$meth
meth["lchl"] <- "~log(chl)"
pred <- ini$pred
pred[c("hyp", "chl"), "lchl"] <- 0
pred["bmi", "chl"] <- 0
pred
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  seed = 38788, print = FALSE)
head(complete(imp))
```



#### Index chunk number 41: transform2
```{r}
nhanes2.ext <- cbind(nhanes2, lchl = NA)
```



#### chunk number 42: index1
```{r}
md.pattern(boys[, c("hgt", "wgt", "bmi")])
```



#### chunk number 43: index2
```{r}
ini <- mice(boys, max = 0, print = FALSE)
meth <- ini$meth
meth["bmi"] <- "~I(wgt/(hgt/100)^2)"
pred <- ini$pred
pred[c("wgt", "hgt", "hc", "reg"), "bmi"] <- 0
pred[c("gen", "phb", "tv"), c("hgt", "wgt", "hc")] <- 0
pred
```



#### chunk number 44: index3
```{r}
imp.idx <- mice(boys, pred = pred, meth = meth, maxit = 20, 
  seed = 9212, print = FALSE)
head(complete(imp.idx)[is.na(boys$bmi), ], 3)
```



#### chunk number 45: index4
```{r}
meth["bmi"] <- "~round(wgt/(hgt/100)^2,dig=2)"
```



#### sum scores chunk number 46: sumscore1
```{r}
ini <- mice(cbind(boys, mat = NA), max = 0, print = FALSE)
meth <- ini$meth
meth["mat"] <- "~I(as.integer(gen) + as.integer(phb) +
 + as.integer(cut(tv,breaks=c(0,3,6,10,15,20,25))))"
meth["bmi"] <- "~I(wgt/(hgt/100)^2)"
pred <- ini$pred
pred[c("bmi", "gen", "phb", "tv"), "mat"] <- 0
pred[c("hgt", "wgt", "hc", "reg"), "mat"] <- 1
pred[c("hgt", "wgt", "hc", "reg"), c("gen", "phb", 
  "tv")] <- 0
pred[c("wgt", "hgt", "hc", "reg"), "bmi"] <- 0
pred[c("gen", "phb", "tv"), c("hgt", "wgt", "hc")] <- 0
pred
```



#### chunk number 47: sumscore2
```{r}
imp.sum <- mice(cbind(boys, mat = NA), pred = pred, 
  meth = meth, maxit = 20, seed = 10948, print = FALSE)
```



#### chunk number 48: sumscore3 
```{r}
xyplot(imp.sum, mat ~ age | .imp, na = gen | phb | 
  tv, subset = .imp == 1, ylab = "Maturation score", 
  xlab = "Age (years)")
```



#### interactions chunk number 50: interaction1
```{r}
nhanes2.ext <- cbind(nhanes2, bmi.chl = NA)
ini <- mice(nhanes2.ext, max = 0, print = FALSE)
meth <- ini$meth
meth["bmi.chl"] <- "~I((bmi-25)*(chl-200))"
pred <- ini$pred
pred[c("bmi", "chl"), "bmi.chl"] <- 0
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  seed = 51600, print = FALSE)
```



#### chunk number 51: interaction2
```{r}
head(ini$pad$data, 3)
```



#### interaction chunk number 52: interaction3
```{r}
nhanes2.ext <- cbind(nhanes2, age.1.bmi = NA, age.2.bmi = NA)
ini <- mice(nhanes2.ext, max = 0, print = FALSE)
meth <- ini$meth
meth["age.1.bmi"] <- "~I(age.1*(bmi-25))"
meth["age.2.bmi"] <- "~I(age.2*(bmi-25))"
pred <- ini$pred
pred[c("age", "bmi"), c("age.1.bmi", "age.2.bmi")] <- 0
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  maxit = 10, print = FALSE)
```



#### squeeze chunk number 53: squeeze1
```{r}
nhanes2.ext <- cbind(nhanes2, lchl = NA)
ini <- mice(nhanes2.ext, max = 0, print = FALSE)
meth <- ini$meth
meth[c("lchl", "chl")] <- c("~log(chl)", "norm")
pred <- ini$pred
pred[c("hyp", "chl"), "lchl"] <- 0
pred["bmi", "chl"] <- 0
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  seed = 1, maxit = 100, print = FALSE)
```



#### chunk number 54: squeeze2 
```{r}
meth["lchl"] <- "~log(squeeze(chl, bounds=c(100,300)))"
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  seed = 1, maxit = 100, print = FALSE)
```



#### post-processing: chunk number 55: post1
```{r}
nhanes2.ext <- cbind(nhanes2, lchl = NA)
ini <- mice(nhanes2.ext, max = 0, print = FALSE)
meth <- ini$meth
meth[c("lchl", "chl")] <- c("~log(chl)", "norm")
pred <- ini$pred
pred[c("hyp", "chl"), "lchl"] <- 0
pred["bmi", "chl"] <- 0
post <- ini$post
post["chl"] <- "imp[[j]][,i] <- squeeze(imp[[j]][,i],c(100,300))"
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  post = post, seed = 30031, maxit = 10, print = FALSE)
imp$imp$chl
```


#### post-processing: chunk number 56: post2
```{r}
ini <- mice(cbind(boys, mat = NA), max = 0, print = FALSE)
meth <- ini$meth
meth["mat"] <- "~I(as.integer(gen) + as.integer(phb) +
 + as.integer(cut(tv,breaks=c(0,3,6,10,15,20,25))))"
meth["bmi"] <- "~I(wgt/(hgt/100)^2)"
pred <- ini$pred
pred[c("bmi", "gen", "phb", "tv"), "mat"] <- 0
pred[c("hgt", "wgt", "hc", "reg"), "mat"] <- 1
pred[c("hgt", "wgt", "hc", "reg"), c("gen", "phb", 
  "tv")] <- 0
pred[c("wgt", "hgt", "hc", "reg"), "bmi"] <- 0
pred[c("gen", "phb", "tv"), c("hgt", "wgt", "hc")] <- 0
pred
```



#### chunk number 57: post3
```{r}
post <- ini$post
post["gen"] <- "imp[[j]][p$data$age[!r[,j]]<5,i] <- levels(boys$gen)[1]"
post["phb"] <- "imp[[j]][p$data$age[!r[,j]]<5,i] <- levels(boys$phb)[1]"
post["tv"] <- "imp[[j]][p$data$age[!r[,j]]<5,i] <- 1"
imp <- mice(cbind(boys, mat = NA), pred = pred, meth = meth, 
  post = post, maxit = 10, print = FALSE)
```



#### chunk number 58: post4 
```{r}
xyplot(imp, mat ~ age | .imp, na = gen | phb | tv, 
  subset = .imp == 1, ylab = "Maturation score", 
  xlab = "Age (years)")
```



### 3.6 Visiting scheme

#### chunk number 60: visit1
```{r}
nhanes2.ext <- cbind(nhanes2, bmi.chl = NA)
ini <- mice(nhanes2.ext, max = 0, print = FALSE)
meth <- ini$meth
meth["bmi.chl"] <- "~I((bmi-25)*(chl-200))"
pred <- ini$pred
pred[c("bmi", "chl"), "bmi.chl"] <- 0
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  seed = 51600, print = FALSE)
```



#### chunk number 61: visit2
```{r}
imp$vis
```



#### chunk number 62: visit3 
```{r}
vis <- imp$vis
vis <- append(vis, vis[4], 1)
vis
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  vis = vis, print = FALSE)
```



#### chunk number 63: visit4 
```{r}
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  vis = c(2, 4, 5, 3), print = FALSE)
```



#### chunk number 64: visit5 
```{r}
imp <- mice(nhanes2.ext, meth = meth, pred = pred, 
  vis = "monotone", print = FALSE)
```



### 4.1 Dry run

#### chunk number 65: dryrun1
```{r}
ini <- mice(nhanes2, maxit = 0)
```



#### chunk number 66: dryrun2
```{r}
import <- matrix(c(30, 30, 30, 29, 25, 21, 25, 25, 
  22, 33, 27, 22, 27, 35, 27, 20, 27, 30), byrow = TRUE, 
  nr = 9)
imp <- mice(nhanes, print = FALSE, seed = 77172)
imp$imp$bmi[, 1:2] <- import
imp$imp$bmi
```



### 4.2 Step by Step

#### chunk number 67: step1
```{r}
imp <- mice(nhanes, maxit = 4, seed = 44612, print = FALSE)
imp1 <- mice(nhanes, maxit = 1, seed = 44612, print = FALSE)
a <- runif(10)
imp2 <- mice.mids(imp1, maxit = 3, print = FALSE)
```



#### chunk number 68: step2
```{r}
all(imp$imp$bmi == imp2$imp$bmi)
```



### 4.3 Assessing convergence

#### first path chunk number 69: convergence1
```{r}
ini <- mice(boys, max = 0, print = FALSE)
meth <- ini$meth
meth["bmi"] <- "~I(wgt/(hgt/100)^2)"
meth["wgt"] <- "~I(bmi*(hgt/100)^2)"
meth["hgt"] <- "~I(100*sqrt(wgt/bmi))"
# this has been outcommented since it does not work anymore
# in more recent versions of R
# imp1 <- mice(boys, meth = meth, maxit = 20, print = FALSE, 
# seed = 9212)
```


#### chunk number 70: convergence1plotcmd 
```{r}
# plot(imp1, c("hgt", "wgt", "bmi"))
```



#### second path chunk number 72: convergence2
```{r}
ini <- mice(boys, max = 0, print = FALSE)
meth <- ini$meth
meth["bmi"] <- "~I(wgt/(hgt/100)^2)"
imp2 <- mice(boys, meth = meth, maxit = 20, print = FALSE, 
  seed = 9212)
```

#### chunk number 73: convergence2plotcmd 
```{r}
plot(imp2, c("hgt", "wgt", "bmi"))
```



#### convergenc chunk number 75: convergence3plotcmd
```{r}
plot(imp.idx, c("hgt", "wgt", "bmi"))
```




#### chunk number 77: convergence4plotcmd 
```{r}
plot(imp.idx, c("hc", "gen", "phb"), tick.number = 3)
```


#### number 79: convergence5
```{r}
m <- 5
T <- 20
imp.kendall <- mice(boys, m = m, meth = imp.idx$meth, 
  pred = imp.idx$pred, maxit = 0, print = FALSE)
tau <- matrix(NA, nrow = T, ncol = m)
for (i in 1:T) {
  if (i == 1) 
    set.seed(9212)
  imp.kendall <- mice.mids(imp.kendall, maxit = 1, 
    print = FALSE)
  x <- complete(imp.kendall, "repeated")[, paste("gen", 
    1:m, sep = ".")]
  y <- complete(imp.kendall, "repeated")[, paste("phb", 
    1:m, sep = ".")]
  xn <- as.data.frame(lapply(x, as.numeric))
  yn <- as.data.frame(lapply(y, as.numeric))
  tau[i, ] <- diag(cor(xn, yn, method = "kendall"))
}
```



#### chunk number 80: convergence5plotcmd
```{r}
matplot(x = 1:T, y = tau, xlab = "Iteration", type = "l")
```



#### chunk number 81: convergence5plot
```{r}
matplot(x = 1:T, y = tau, xlab = "Iteration", type = "l")
```


### 4.5 Checking your imputations

#### density plot chunk number 82: densityplotcmd
```{r}
densityplot(imp.kendall, scales = list(x = list(relation = "free")), 
  layout = c(5, 1))
```



#### chunk number 84: propensity
```{r}
hc.na <- is.na(boys$hc)
fit.hc <- with(imp.kendall, glm(hc.na ~ age + wgt + 
  hgt + reg, family = binomial))
ps <- rep(rowMeans(sapply(fit.hc$analyses, fitted.values)), 
  6)
```



#### chunk number 85: propensityplotcmd 
```{r}
xyplot(imp.kendall, hc ~ ps | .imp, pch = c(1, 20), 
  cex = c(0.8, 1.2), xlab = "Probability that head circumference is missing", 
  ylab = "Head circumference (cm)", scales = list(tick.number = 3))
```



#### calculate model and residuals chunk number 87: residuals1
```{r}
hc <- complete(imp.kendall, "long", TRUE)$hc
fit <- lm(hc ~ poly(ps, 4))
```



#### chunk number 88: residuals2 
```{r}
densityplot(~residuals(fit), group = hc.na, plot.points = FALSE, 
  ref = TRUE, scales = list(y = list(draw = FALSE)), 
  par.settings = simpleTheme(col.line = rep(mdc(1:2))), 
  xlab = "Residuals of regression of hc on propensity score", 
  lwd = 2)
```



### 5.1 Repeated data analysis

#### chunk number 90: repeated1
```{r}
imp <- mice(nhanes2, seed = 99210, print = FALSE)
fit <- with(imp, lm(chl ~ age + bmi))
summary(pool(fit))
```



#### chunk number 91: repeated2
```{r}
expr <- expression(ov <- cut(bmi, c(10, 25, 50)), table(age, 
  ov))
fit <- with(imp, eval(expr))
```



#### chunk number 92: repeated3
```{r}
fit$an[c(2, 5)]
```



#### chunk number 93: extract1
```{r}
com <- complete(imp, 3)
```



#### chunk number 94: extract2
```{r}
com <- complete(imp, "long")
```



#### chunk number 95: extract3
```{r}
com <- complete(imp, "long", include = TRUE)
by(cbind(age = com$age, ov = cut(com$bmi, c(10, 25, 
  50))), com$.imp, table)
```



### 6.1 Adding your own imputation functions

#### chunk number 96: pool1
```{r}
fit <- with(imp, lm(chl ~ age + bmi))
est <- pool(fit)
```



#### chunk number 97: pool2
```{r}
methods(coef)
methods(vcov)
```


#### Model testing - Wald method chunk number 98: pool3
```{r}
imp <- mice(nhanes2, print = FALSE, m = 50, seed = 219)
fit0 <- with(data = imp, expr = lm(bmi ~ age + hyp))
fit1 <- with(data = imp, expr = lm(bmi ~ age + hyp + 
  chl))
stat <- pool.compare(fit1, fit0, method = "Wald")
stat$p
```



#### Model testing - LLR method chunk number 99: pool4
```{r}
imp <- mice(boys, print = FALSE, seed = 60019)
fit0 <- with(data = imp, expr = glm(I(gen > levels(gen)[1]) ~ 
  hgt + hc, family = binomial))
fit1 <- with(data = imp, expr = glm(I(gen > levels(gen)[1]) ~ 
  hgt + hc + reg, family = binomial))
stat <- pool.compare(fit1, fit0, method = "likelihood", 
  data = imp)
stat$p
```




#### chunk number 100: ownfunction 
```{r}
mice.impute.myfunc <- mice.impute.pmm
# make some edits of myfunc, add run your method as
mice(nhanes, method = "myfunc", print = FALSE)
```


### 6.2 Sensitivity analysis under MNAR

#### chunk number 101: sensitivity
```{r}
ini <- mice(nhanes2, maxit = 0, print = FALSE)
post <- ini$post
k <- seq(1, 1.5, 0.1)
est <- vector("list", length(k))
for (i in 1:length(k)) {
  post["chl"] <- paste("imp[[j]][,i] <-", k[i], "* imp[[j]][,i]")
  imp <- mice(nhanes2, post = post, seed = 10, print = FALSE, 
    maxit = 20)
  fit <- with(imp, lm(bmi ~ age + chl))
  est[[i]] <- summary(pool(fit))
}
```



### 7.2 SPSS
```{r}
# The code below should not be run from R, but from
# SPSS.  Remove one (1) # sign per line, and run.
# BEGIN PROGRAM R.  Instruct R to import the data
# and the dictionary from SPSS.  dict <-
# spssdictionary.GetDictionaryFromSPSS() data <-
# spssdata.GetDataFromSPSS() Load mice, impute and
# get the original + imputed data. library(mice)
# imp <- mice(data,maxit=10) com <-
# complete(imp,'long',inc=TRUE) com <- cbind(com,
# Imputation_ = as.integer(com$.imp)-1) Export
# imputed data and dictionary from R to SPSS.
# spssdictionary.SetDictionaryToSPSS('com', dict)
# spssdata.SetDataToSPSS('com', com)
# spssdictionary.EndDataStep() END PROGRAM.
```

### 7.3 MItools

#### chunk number 102: mitools1
```{r}
mydata <- mitools::imputationList(lapply(1:5, complete, x = imp))
```



#### chunk number 103: mitools2 
```{r}
fit <- with(mydata, expr = lm(chl ~ age + bmi))
```


### 7.3 Zelig

#### chunk number 104: zelig1
```{r}
library("Zelig", quietly = TRUE)
# NOTE: we must now call mice:::cbind.mids explicitly because cbind() does  
# not use the standard S3 dispatch mechanism. Since April 2017 it 
# dispatches to cbind.data.frame() when one of the arguments is a 
# data frame. See also https://github.com/tidyverse/dplyr/issues/2667
imp <- mice:::cbind.mids(imp.idx, data.frame(r.hc = is.na(boys$hc)))
mydata2 <- mi(complete(imp, 1), complete(imp, 2), complete(imp, 
  3), complete(imp, 4), complete(imp, 5))
fit <- zelig(r.hc ~ age + wgt + hgt + bmi + gen + phb + 
  tv + reg, model = "logit", data = mydata2, cite = FALSE)
summary(fit)
```

