---
title: "Help for old friends"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Help for old friends}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<!-- oldfriends.md is generated from oldfriends.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Help for old friends

This documents describes the changes between `mice 2.46.0` and 
`mice 3.0.0`. Most code written for versions `mice 2.12 - mice 2.46.0`
should run unchanged. I have tried to minimize the changes to the 
function arguments, but it was not possible to remain 100% backward 
compatible. This document outlines the visible changes, and suggests
ways how to adapt old code to `mice 3.0`.

## `mice` function arguments

| Argument  | 2.46.0                   | 3.0.0
|:----------|:-------------------------|:-------------------------|
method      | length `ncol(data)`      | length `length(blocks)`  |
predictorMatrix | `ncol(data)` rows and columns | `length(blocks)` rows, `ncol(data)` columns |
blocks      | not defined | new argument |
visitSequence | `integer`, arbitrary length | `character`, arbitrary length |
form        | character, length `ncol(data)`| deprecated |
formulas    | not defined | named list of formulas |
blots       | not defined | named `list` of `alist` |  
diagnostics | | deprecated |
imputationMethod | | deprecated, use `method` |
defaultImputationMethod | | deprecated, use `defaultMethod` |
... | | |

No changes were made for the following arguments: `data`, `m`,
`where`, `post`, `defaultMethod`, `maxit`, `printFlag`, `seed` and
`data.init`. 

If `blocks` is not specified, then each variable is allocated to a
separate block. In that case, `length(blocks)` is identical to
`ncol(data)`, and the method in `mice 3.0.0` reduces to
variable-by-variable imputation, as in `mice 2.46.0` and before.

Argument `visitSequence` may still be specified as `integer` 
or `numeric`, but it will internally be converted into `character` 
using the column names in `data`.

An existing function call to `mice` using the old `form` argument
may result in an error `Argument "formulas" not a list`. The advice
is to specify the formula as a list, e.g.,

```{r}
library(mice, warn.conflicts = FALSE)
imp <- mice(nhanes,
  formulas = list(
    hyp ~ bmi,
    chl ~ age + hyp + bmi,
    bmi ~ age + hyp + chl
  ),
  print = FALSE, m = 1, maxit = 1, seed = 1
)
imp$formulas
```


## `pool()` function uses `broom`

`mice 2.46.0` and before used `coef()` and `vcov()` to extract the parameters
from the complete-data model. There were two problems with this approach: 
1) `coef()` or `vcov()` were often not defined for the analysis of interest, 
2) their output was not standard across procedures. Older versions of 
`mice` therefore needed quite some custom code to extract the parameters 
and their variance estimates.

`mice 3.0.0` uses the `broom` package for this task. The advantage is that the
standardised output. The downside is that (still) many packages offer no 
defined `tidy.xxx()` and `glance.xxx()` functions. In those cases, the user
sees the error message `Error: No tidy methods for objects of class xxx`. See 
the pool() documentation for what to do in those cases.

## `pool()` for mixed models requires `library(broom.mixed)`

`mice` automatically loads the `broom` package. Tidiers for mixed models live 
in the `broom.mixed` packages and this is NOT automatically loaded. If you want 
to pool the results of mixed model, issue `library(broom.mixed)` before calling
the `pool()` function.
