---
title: "Random effects strategies in `mice.impute.2l.bin`"
author: "Stef van Buuren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random effects strategies in `mice.impute.2l.bin`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Background

`mice.impute.2l.bin()` imputes a binary outcome in a two-level (clustered)
data structure using a logistic mixed model fitted by `lme4::glmer()`. A key
modelling choice is how the cluster-specific random effect $b_i^*$ is drawn
for clusters that contain missing values.

Audigier et al. (2018, p. 163) describe two situations for the FCS-GLM
imputer:

- **Systematic missingness**: cluster $i$ has *no* observed values of $y$.
  Nothing is known about $b_i$ from the data, so the only coherent draw is
  from the marginal prior $b_i^* \sim N(0, \psi^*)$.

- **Sporadic missingness**: cluster $i$ has *some* observed values of $y$.
  These observations are informative about $b_i$, so the draw should come
  from the posterior $p(b_i \mid y_{\text{obs},i}, X_i, \beta^*, \psi^*)$.

`mice.impute.2l.bin()` exposes three strategies via the `random.effects`
argument.

## The three strategies

### `"marginal"` — original behaviour

$$b_i^* \sim N(0, \psi^*)$$

All missing clusters, sporadic or systematic, draw from the marginal prior.
This ignores the observed responses within a partially observed cluster.

**Pro:** Simple; correct for systematic missingness.
**Con:** For sporadic clusters, ignores information in $y_{\text{obs},i}$,
leading to overdispersed imputations.

### `"eb"` — Empirical Bayes

$$b_i^* \sim N(\hat{b}_i, \psi^*)$$

The draw is centred on the BLUP $\hat{b}_i$ from `lme4::ranef()`, but still
spread with the full marginal variance $\psi^*$.

**Pro:** Uses the observed data to locate $b_i^*$; easy to implement.
**Con:** Does not account for the uncertainty in $\hat{b}_i$. Clusters with
few observations get the same spread as clusters with many observations,
underestimating uncertainty for small clusters and overestimating it for
large ones.

### `"laplace"` — Laplace approximation (default)

$$b_i^* \sim N(m_i, V_i)$$

where $m_i$ is the BLUP and $V_i$ is the Laplace-approximated posterior
variance returned by `lme4::ranef(fit, condVar = TRUE)`.

As $n_i \to 0$, $V_i \to \psi^*$ and $m_i \to 0$, recovering the marginal
draw. As $n_i \to \infty$, $V_i \to 0$ and $m_i \to \hat{b}_i$,
concentrating the draw around the BLUP.

**Pro:** Posterior uncertainty in $b_i$ is correctly reflected; bridges
sporadic and systematic cases continuously; consistent with the
recommendation of Audigier et al. (2018).
**Con:** Relies on the Laplace approximation, which can be inaccurate for
very sparse binary data (few events per cluster).

## Simulation

We simulate a two-level dataset with known parameters to make the differences
between strategies visible. Strong random effects ($\sigma_b = 3$) and a mix
of sporadic and systematic missingness create conditions where the three
methods behave differently.

```{r simulate, message=FALSE, warning=FALSE}
library(mice)
library(ggplot2)

set.seed(42)

# simulate data
n_clusters    <- 60
n_per_cluster <- 8
sigma_b       <- 3
beta0         <- -0.5
beta1         <-  1.0

cluster <- rep(seq_len(n_clusters), each = n_per_cluster)
x1      <- rnorm(n_clusters * n_per_cluster)
b_true  <- rnorm(n_clusters, 0, sigma_b)
logit   <- beta0 + beta1 * x1 + b_true[cluster]
y       <- rbinom(length(logit), 1, plogis(logit))
dat     <- data.frame(cluster = cluster, x1 = x1, y = y)

# systematic missingness: 30% of clusters have all y missing
sys_cl     <- sample(seq_len(n_clusters), 18)
dat$y_miss <- dat$y
dat$y_miss[dat$cluster %in% sys_cl] <- NA

# sporadic missingness: 60% of remaining observations missing
spor_idx <- which(!is.na(dat$y_miss))
dat$y_miss[sample(spor_idx, round(0.6 * length(spor_idx)))] <- NA

n_obs_per_cluster <- tapply(!is.na(dat$y_miss), dat$cluster, sum)
cat(sprintf(
  "Clusters: %d total, %d systematic (all missing), %d sporadic\n",
  n_clusters, length(sys_cl), n_clusters - length(sys_cl)
))
cat(sprintf(
  "Observed y per sporadic cluster: median %g, range %g-%g\n",
  median(n_obs_per_cluster[n_obs_per_cluster > 0]),
  min(n_obs_per_cluster[n_obs_per_cluster > 0]),
  max(n_obs_per_cluster[n_obs_per_cluster > 0])
))
cat("True proportion:", round(mean(y), 3), "\n")
```

```{r impute, message=FALSE, warning=FALSE, results="hide"}
pred <- make.predictorMatrix(dat[, c("cluster", "x1", "y_miss")])
pred["y_miss", "cluster"] <- -2
pred["y_miss", "x1"]      <-  1
pred[c("cluster", "x1"), ] <- 0

m     <- 40
maxit <-  2

run_imp <- function(re) suppressWarnings(
  mice(dat[, c("cluster", "x1", "y_miss")],
       method = c("", "", "2l.bin"),
       pred = pred, m = m, maxit = maxit,
       print = FALSE, seed = 1, random.effects = re)
)

imp_marginal <- run_imp("marginal")
imp_eb       <- run_imp("eb")
imp_laplace  <- run_imp("laplace")
```

### Between-imputation variability per cluster

The key difference between methods appears at the cluster level: how much
do the imputed cluster means vary across the $m$ imputations?

```{r cluster-var, warning=FALSE}
get_cluster_sd <- function(imp, label) {
  vals <- sapply(seq_len(imp$m), function(i) {
    tapply(complete(imp, i)$y_miss, complete(imp, i)$cluster, mean)
  })
  data.frame(
    method    = label,
    n_obs     = as.numeric(n_obs_per_cluster),
    cluster_sd = apply(vals, 1, sd)
  )
}

csds <- rbind(
  get_cluster_sd(imp_marginal, "marginal"),
  get_cluster_sd(imp_eb,       "eb"),
  get_cluster_sd(imp_laplace,  "laplace")
)

ggplot(csds, aes(x = n_obs, y = cluster_sd, colour = method)) +
  geom_jitter(alpha = 0.4, width = 0.15) +
  geom_smooth(se = FALSE, method = "loess", span = 1.2) +
  labs(x = "Observed y in cluster", y = "SD of imputed cluster mean",
       title = "Between-imputation variability per cluster") +
  theme_bw()
```

For clusters with zero observed $y$ (systematic, $n_\text{obs} = 0$), all
three methods draw from the marginal prior and are identical. As
$n_\text{obs}$ increases:

- `"marginal"` stays high — it ignores the observed data within the cluster
  and always draws from the prior.
- `"eb"` drops — it centres on the BLUP, but uses the full prior width, so
  variability stays moderate.
- `"laplace"` drops fastest — it uses both the BLUP and the tighter
  posterior variance, giving the most precise draws for well-observed
  clusters.

### Cluster-level bias

```{r cluster-bias, warning=FALSE}
true_cluster_p <- tapply(plogis(logit), cluster, mean)

get_bias <- function(imp, label) {
  mean_imp <- rowMeans(sapply(seq_len(imp$m), function(i) {
    tapply(complete(imp, i)$y_miss, complete(imp, i)$cluster, mean)
  }))
  data.frame(
    method   = label,
    true_p   = true_cluster_p,
    imp_mean = mean_imp,
    n_obs    = as.numeric(n_obs_per_cluster)
  )
}

bias <- rbind(
  get_bias(imp_marginal, "marginal"),
  get_bias(imp_eb,       "eb"),
  get_bias(imp_laplace,  "laplace")
)

ggplot(bias, aes(x = true_p, y = imp_mean, colour = n_obs)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~method) +
  scale_colour_viridis_c(name = "n obs") +
  labs(x = "True cluster probability", y = "Mean imputed cluster probability",
       title = "Imputed vs true cluster probability") +
  theme_bw()
```

`"laplace"` and `"eb"` track the true cluster probabilities more closely
than `"marginal"`, particularly for extreme clusters (high or low true
probability). `"marginal"` shows shrinkage toward the population mean because
it ignores the cluster's own observed data.

## Recommendation

Use `random.effects = "laplace"` (the default) for most applications. It is
the only method that correctly handles the full spectrum from sporadic to
systematic missingness within a single coherent framework.

Use `random.effects = "marginal"` only if all missing clusters are known to
be systematically missing (no observed $y$ in the cluster at all).

Use `random.effects = "eb"` as a fast approximation when clusters are large
(many observations per cluster), so that posterior uncertainty in $\hat{b}_i$
is small relative to $\psi^*$.

## References

Audigier, V., White, I. R., Jolani, S., Debray, T. P. A., Quartagno, M.,
Carpenter, J., van Buuren, S., & Resche-Rigon, M. (2018). Multiple imputation
for multilevel data with continuous and binary variables. *Statistical
Science*, 33(2), 160–183. https://doi.org/10.1214/18-STS646
